Informing policy via dynamic models: Cholera in Haiti

### 2024-02-15 r1/response.tex

\report{
  My most substantial concern is with the framing of the paper, and whether that might mislead potential readers. To me, the paper serves more of an educational purpose, presenting ideas on how to best use mathematical models to gain insight from time series data. While the case study into cholera in Haiti is methodologically sound, it does not seem to be the main focus of the manuscript. To such an extent that, in the discussion, the only comment to on the Haiti analysis is that the authors ``used the same data and models, and even much of the same code, as Lee et al. [1], and yet ended up with drastically different conclusions."
}

\report{
  This major concern can be resolved in two ways, both of which seem acceptable to me, either a) reframe the paper as a more general series of recommendations for how to better perform statistical inference on time series data using mechanistic models (including a longer introduction and discussion of existing methodological deficiencies) or b) perform a more in-depth analysis and commentary on cholera in Haiti (some of which has already been done but is buried in supplementary information files that are only briefly mentioned in the main text, e.g. S5).
}

Thanks for the feedback on the framing of our contribution.
The original goal of our manuscript was to combine the two foci identified by the referee:
we use an in-depth case study to motivate general procedures for fitting mathematical models to time series data, while explaining how these procedures help to address limitations of previous approaches.
The case study was intended both to help motivate the advice and to make sure that the advice is pertinent and practical on at least one suite of models.
Since our approach is predicated on the possibility of carrying out some nontrivial computational procedures, we are led to address computational issues in parallel with scientific and statistical issues.

The referee would prefer us to emphasize one of these foci, and this feedback has helped us see how readers might find themselves struggling to orient themselves while reading the manuscript.
We agree that the ``educational'' purpose of the paper is primary to the specific conclusions about the Haiti cholera epidemic.
However, the validity of our reanalysis of the Haiti data and models provides critical support to our methodological points.
For the revision, we have edited the abstract and introduction to make explicit how the case study fits into the overall goals of the paper.
In the discussion, we return from the concrete consideration of the case study, placing it back in the more general setting of inferring complex dynamics from time series data.
In the revision, we have added more signposts to help the reader see how the points that arise in the discussion generalize the corresponding points arising in the case study.

@@

Thank you for this feedback.
We have made substantial changes to the abstract, author summary, and introduction in order to make the focus of the paper more clear.
We have also done a nearly complete rewritting of the discussion section.
Specifically, though the content and arguments made remain the same, we specifically point out ways that our case study demonstrates the usefulness of our general suggestions for model based inference of dynamic systems.

\report{
  \textbf{\underline{Abstract}}

  The authors state they ``develop data analysis strategies leading to improved statistical fit." It's not clear to me exactly what novel strategies have been developed, it seems the authors exclusively use pre-existing techniques from the literature. This is fine, if the focus of the paper is either a) pedagogical or b) on understanding a specific disease system (cholera in Haiti). If any methods are novel, this should be more clearly stated.
}

Fitting models 1 and 2 does not require any methodological innovation. Model 3 has a stochastic spatiotemporal partially-observed structure that leads to methodological challenges. \cite{lee20} made some approximations to avoid carrying out inference for the full coupled system. We show some consequences of those approximations, taking advantage of a newly developed method which has not yet been applied in the context of scientific investigations. So, although the method is not entirely new, its applicability to this task is novel.
Additional commentary to this point has been added in the abstract, introduction, and the subsection on fitting Model 3.

@@

\article{We assess current methodological approaches to these issues via a case study of the 2010-2019 cholera epidemic in Haiti. We consider three dynamic models developed by expert teams to advise on vaccination policies. We evaluate previous methods used for fitting these models, and we demonstrate modified data analysis strategies leading to improved statistical fit. Speficially, we present appraoches of diagnosing model misspecification and the consequent development of improved models. We futher demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for high-dimensional models for spatiotemporal incidence data.}

\report{
  \textbf{\underline{Introduction}}

1.  To make clear that the issues being addressed by this paper go beyond one modelling study of Haiti, the authors should provide examples and references to support the (accurate in my opinion) statement about, ``common modeling decisions that may not provide an adequate statistical explanation of the data".
}

That statement arose in the author summary, where references are not expected.
However, we appreciate that the point should be expanded on, and we have done so in the revision. Specifically, we cite papers from the literature review we performed on the use of dynamic models for analyzing the Haiti cholera oubreak. The literature review was previously mentioned, but citations from the review were not made. Furthermore, we modified the quoted statement in the author summary to be the following:

@@

\article{we use an in-depth case study to motivate general procedures for fitting mathematical models to time series data, while explaining how these procedures help to address limitations of previous approaches.}

Among the most common mistakes in mechanistic modeling, in our opinion, is not properly assessing / evaluating a fitted model.
We originally tried to make this point by having an entire section dedicated to model diagnostics, though we admit that the connection between the statement in the author summary and a section later in the article was not clear.
\TODO{We have addressed this by adding sign-posts to references and suggestions back to this statement, and other similar statements in the abtract / introduction.}
Our modified discussion also includes detailed examples of principles that were very useful in our case study but are not universal practice.

@@

The referee raises several points in this paragraph, and we consider them one at a time.

We agree that adding to model complexity has costs, including loss of identifability.
AIC and other formal statistical methods can assess whether the additional complexity is justified by a sufficiently large improvement of statistical fit. Identifiability and model simplicity have practical scientific value which goes beyond their statistical role in reducing generalization error and hence improving out-of-fit predictions. This is a variation on the issue of ``statistical significance versus practical significance" familiar from introductory statistics textbooks.

Moving from an ODE model to a stochastic model may not involve any additional parameters, if only demographic stochasticity is included.
Generally, it is advisable to consider the possibility of over-dispersed variation on the transmission rate \cite{breto09,he10} leading to one extra parameter compared to a deterministic model.
In the context of cholera dynamics, \cite{lemaitre19} found a substantial improvement in model fit by including stochasticity, involving one over-dispersion parameter.
\cite{stocks20} found similar results, favoring a stochastic dynamic model with over-dispersion, when investigating transmission dynamics of a different enteric pathogen, rotavirus.
In other contexts, \cite{king15} noted various scientific advantages for including stochasticity in the model; it is undoubtedly part of the biological dynamics, and its omission can lead to over-confident predictions.

In our study, we find that the deterministic model (model 2) has poor statistical fit.
This can lead to a risk of identifying spurious scientific phenomena: additional deterministic complexity added to the model may help to describe unmodeled stochastic phenomena present in the data, and may therefore spuriously be assessed as a highly statistically significant model improvement.
We see some evidence of this in our analysis.
Aside from not being stochastic, the mechanistic structure of Model~2 is far more complex than the other two models considered in our study; we also see the most clear cases of spurious scientific findings, as some estimated parameters favor regions of the parameter space that we would not expect scientifically (see the added supplement material for confidence intervals of model parameters).
This may partially explain the choice of Lee et al~\cite{lee20} to consider a scientifically unjustifiable jump in latent state values from their epidemic to endemic periods: without this additional flexibility, Model~2 cannot adequately describe cholera dynamics in Haiti while keeping parameter values in scientifically plausible ranges.
From our experience, this type of a result is quite common: for a deterministic model to adequately account for unmodeled features of a dynamic system, parameters that are estimated are often forced to unexpected regions of the parameter space.
Adding stochasticity to the model helps address this issue by capturing unmodeled features in the inherent stochasticity of the model rather than in estimated parameters.
Stochasticity in the model alone may not fully solve this problem as evidence suggests that the failure to account for extra-demographic variablility may lead to biases in parameter estimates \cite{he10}.
Our profile likelihood results for this case study also provide strong evidence for the inclusion of extra-demographic stochasticity in Models~1 and 3 (see supplement material section S8).

The decision on whether to choose Markov chain Monte Carlo (MCMC) methods is not fully aligned with the decision on whether to include stochasticity in the model.
In our literature review, of the 12 previous Haiti cholera investigations including stochastic dynamic models \cite{kirpich17,lee20,pasetto18,mukandavire13,kirpich16,lewnard16,kunkel17,mukandavire15,sallah17,azman12,azman15}, four of them carried out inference via MCMC \cite{pasetto18,lewnard16,sallah17,azman12}.
We agree with the referee that MCMC is more routine for determinstic models than for stochastic models, and advanced MCMC methods such as Hamiltonian Monte Carlo become more readily applicable.
Sequential Monte Carlo (SMC) was used for only two of the investigations in our literature review \cite{lee20,azman15}.
In principle, MCMC methods can be applied for inference in over-dispersed stochastic dynamic models via so-called particle MCMC (PMCMC) \cite{andrieu10}.
In practice, these models have been fitted by maximum likelihood, some examples including \cite{lemaitre19,fox22,molodecky23,pons-salort18,subramanian20,stocks20}.
In this case, identifiability is generally addressed via profile likelihood methods.

### 2024-02-15 edits.tex
Addressed several minor TODO items

\newcommand\editMechModels{Mechanistic models representing biological phenomena are valuable for epidemiology and consequently for public health policy \cite{lofgren14,mccabe21}. More broadly, they have useful roles throughout biology, expecially when combined with statistical methods that properly account for stochasticity and nonlinearity \cite{may04}. In some situations, modern machine learning methods, which are associative by nature, can outperform mechanistic models on epidemiological forecasting tasks \cite{lau22,baker18}. The predictive skill of non-mechansitic models can reveal limitations in mechanistic models, but cannot readily replace the scientific understanding obtained by describing the biological dynamics of the system in a mathematical model \cite{baker18}.}

@@

\newcommand\editDiscussionFour{The three models in this case study offer a unique combination of decisions of model complexity. Model~3 provides an interesting example in that it is both stochastic and has a meta-population structure, making it challenging to draw likelihood-based inferences. In this paper, we demonstrated how this model class could be calibrated to incidence data using the innovative IBPF algorithm. One of only a few examples of fitting a model with this level of complexity, this case study exemplifies the algorithm's potential benefits and provides an example for future researchers on a possible approach to fitting a high-dimensional non-linear model.}

\newcommand\editDiscussionFive{If forecasts are an important component of a modeling task, the most recent information on the dynamic system should be considered with increased importance. This is particularly true for long time series, as the latent states from simulations from initial conditions may be more likely to diverge from the truth as time increases. One way to address this issue is by simulating forward from the filtering distribution at the last available time point, as this proceedure considers available evidence about the latent states of a system under the framework of the dynamic model. This type of forecasting, however, is not possible using a deterministic model. Deterministic simulations may also result in over-confidence in model forecasts, as they can only account for uncertainty due to the parameter estimation proceedure \cite{king15}. Despite these limitations, forecasts from deterministic models can provide valuable insights on potential future trends, though researchers should be transparent about these limitations when presenting forecasts from deterministic models.}

>>>>

\newcommand\editMechModels{Mechanistic models representing biological phenomena are valuable for epidemiology and consequently for public health policy \cite{lofgren14,mccabe21}. More broadly, they have useful roles throughout biology, expecially when combined with statistical methods that properly account for stochasticity and nonlinearity \cite{may04}. In some situations, modern machine learning methods can outperform mechanistic models on epidemiological forecasting tasks \cite{lau22,baker18}. The predictive skill of non-mechansitic models can reveal limitations in mechanistic models, but cannot readily replace the scientific understanding obtained by describing the biological dynamics of the system in a mathematical model \cite{baker18}.}

@@

\newcommand\editDiscussionFour{Unmodeled features of a dynamic system can lead to spurious or misleading parameter estimates if the feature greatly impacts observed data. This is because the model must compensate for these unaccounted-for features using its inherent flexibility, particularly in the calibration of parameters. In deterministic models, parameter estimation is the sole source of such flexibility. When the influence of unmodeled dynamics is significant, it can skew parameter estimates toward scientifically implausible values, a phenomenon observed with Model~2 in our case study. Incorporating demographic and environmental stochasticity into models can mitigate the impact of these unmodeled features. These stochastic processes are not only arguably present in most dynamic systems, but their inclusion in a model also allows observed data variations to be attributed to inherent randomness rather than to distorted parameter values. Furthermore, incorporating extra-demographic stochasticity in a model has also been shown to be beneficial \cite{he10,stocks20,li23}, which aligns with our findings: Models~1 and 3 suggest the presence of extra-demographic stochsticity, as evidenced by the confidence intervals provided in Suplement S8.}

\newcommand\editDiscussionFive{The three models in this case study offer a unique combination of decisions of model complexity. Model~3 provides an interesting example in that it is both stochastic and has a meta-population structure, making it challenging to draw likelihood-based inferences. In this paper, we demonstrated how this model class could be calibrated to incidence data using the innovative IBPF algorithm. One of only a few examples of fitting a model with this level of complexity, this case study exemplifies the algorithm's potential benefits and provides an example for future researchers on a possible approach to fitting a high-dimensional non-linear model.}

\newcommand\editDiscussionSix{If forecasts are an important component of a modeling task, the most recent information on the dynamic system should be considered with increased importance. This is particularly true for long time series, as the latent states from simulations from initial conditions may be more likely to diverge from the truth as time increases. One way to address this issue is by simulating forward from the filtering distribution at the last available time point, as this proceedure considers available evidence about the latent states of a system under the framework of the dynamic model. This type of forecasting, however, is not possible using a deterministic model. Deterministic simulations may also result in over-confidence in model forecasts, as they can only account for uncertainty due to the parameter estimation proceedure \cite{king15}. Despite these limitations, forecasts from deterministic models can provide valuable insights on potential future trends, though researchers should be transparent about these limitations when presenting forecasts from deterministic models.}

### 2024-02-15 ms.Rnw
Addressed several minor TODO items

The four independent teams were given the task of estimating the
potential effect of prospective oral cholera vaccine (OCV) programs.
While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \cite{pezzoli20}.
Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \cite{lee20}.
\new{The possibility of controling the Haiti cholera outbreak via OCV was considered by various research groups.} \TODO{Citations}. 
In the \new{Lee et al.~\cite{lee20}} study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
\old{The groups largely adhered to existing guidelines for creating models to inform policy [35, 36] and, d}\new{D}espite their autonomy, \new{the four independent teams} obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti. 
\new{They also provided a range for the forecasted number of cumulative cases from February 2019 to Februaruy 2024 based on model simulations, with median estimates from each model measured in hundreds of thousands.}
\old{Their forecasts, however, were} \new{These forecasts are} inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
Though cholera has recently reemerged in Haiti \cite{rubin22,PAHO23}, the inability to accurately forecast cholera incidence from 2019-2022\old{---while diligently adhering to existing scientific standards for infectious disease modeling---demonstrates that existing standards of disease modeling and forecasting may be insufficient.} \new{prompts us to consider retrospectively what may have been done differently in order to obtain more reliable conclusions.}

@@

# TODO: update this comment: This is used later in figure 2 (TODO)
beta_t_mcap_results <- pomp::mcap(h1_prof_res$logLik, h1_prof_res$betat, span = 1)

@@

  # Otherwise (missing RL3 file, and the corresponding <RUN_LEVEL> file):
  # Throw an error, letting users know how to create the file.
  stop(paste0("Missing file: `model3/", rl_dir, 'h3_MLE_eval.rds`;\nThe file can be obtained by running the script `H3_MLE_Eval.R`.'))
}

# TODO: Remove
# haiti3_fit contains ll estimate for many parameter sets, so we want to find
# the highest value
# mod3_ll <- h3_prof %>%
#   slice_max(order_by = logLik, n = 1) %>%
#   pull(logLik)
@

@@

# model 3 AIC
# TODO: Remove
# mod3_aic <- 2 * mod3_n_params - 2 * mod3_ll

@@

\section*{Materials and methods}
\subsection*{Mechanistic models for disease modeling}
\new{In this article, we refer to} models that focus on learning relationships between variables in a dataset \new{as} \old{are called} {\it associative}, whereas models that incorporate a known scientific property of the system \old{are called} \new{we call} {\it causal} or {\it mechanistic}.
The danger in using forecasting techniques which rely on associative models to predict the consequence of interventions is called the Lucas critique in an econometric context.
Lucas et al.~\cite{lucas76} pointed out that it is naive to predict the effects of an intervention on a given system based entirely on historical associations.
To successfully predict the effect of an intervention, a model should therefore both provide a quantitative explanation of existing data and should have a causal interpretation: a manipulation of the system should correspond quantitatively with the corresponding change to the model.
This motivates the development of mechanistic models, which provides a statistical fit to the available data while also supporting a causal interpretation.
\new{Despite their limited ability to project the effect of interventions on a system, associative models can be effectively used to make inference on an certain features of a system. 
In our literature review, there were 14 studies that used associative models to describe various aspects of the cholera epidemic \cite{hulland19,moise20,piarroux11,matias17,eisenberg13,rebaudet19CATI,ivers15,raila17,cuneo17,charles14,richterman19,richterman19inverse,bengtsson15,li16adjusting}.}

@@

@@

We implemented Model~1 using the \code{pomp} package \cite{king16}, relying heavily on the \code{pomp} source code provided by Lee et al.~\cite{lee20}.
Both analyses used the \code{mif2} implementation of the IF2 algorithm to estimate $\paramVec$ by maximum likelihood.
One change we made in the statistical analysis that led to larger model likelihoods was increasing the computational effort in the numerical maximization.
While IF2 enables parameter estimation for a large class of models, the theoretic ability to maximize the likelihood depends on asymptotics in both the number of particles and the number of filtering iterations.
Many Monte Carlo replications are then required to quantify and further reduce the error.
The large increase in the log-likelihood for Model~1 (see Table~\ref{tab:likes}) can primarily be attributed to increasing the computational effort in fitting the model; this result highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
Simulations from the initial conditions of our fitted model are plotted against the observed incidence data in Fig.~\ref{fig:mod1fit}.

@@

Another approach is to compare a quantitative measure of the model fit (such as MSE, predictive accuracy, or model likelihood) among all proposed models.
These comparisons, which provide insight into how each model performs relative to the others, are quite common \cite{rinaldo12,sallah17}.
\TODO{Add citations from lit-review.}
To calibrate relative measures of fit, it is useful to compare against a model that has well-understood statistical ability to fit data, and we call this model a {\it benchmark}.
Standard statistical models, interpreted as associative models without requiring any mechanistic interpretation of their parameters, provide suitable benchmarks.
Examples include linear regression, auto regressive moving average time series models, or even independent and identically distributed measurements.
Benchmarks enable us to evaluate the goodness of fit that can be expected of a suitable mechanistic model.

@@

>>>>

The four independent teams were given the task of estimating the
potential effect of prospective oral cholera vaccine (OCV) programs.
While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \cite{pezzoli20}.
Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \cite{lee20}.
\new{The possibility of controlling the Haiti cholera outbreak via OCV was considered by various research groups \cite{ivers17,andrews11,walton11,matias17,chao11,fung13,date11,leung22,azman15}}.
In the \new{Lee et al.~\cite{lee20}} study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
\old{The groups largely adhered to existing guidelines for creating models to inform policy [35, 36] and, d}\new{D}espite their autonomy, \new{the four independent teams} obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
\new{They also provided a range for the forecasted number of cumulative cases from February 2019 to Februaruy 2024 based on model simulations, with median estimates from each model measured in hundreds of thousands.}
\old{Their forecasts, however, were} \new{These forecasts are} inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
Though cholera has recently reemerged in Haiti \cite{rubin22,PAHO23}, the inability to accurately forecast cholera incidence from 2019-2022\old{---while diligently adhering to existing scientific standards for infectious disease modeling---demonstrates that existing standards of disease modeling and forecasting may be insufficient.} \new{prompts us to consider retrospectively what may have been done differently in order to obtain more reliable conclusions.}

@@

# This is used later in figure 2
beta_t_mcap_results <- pomp::mcap(h1_prof_res$logLik, h1_prof_res$betat, span = 1)

@@

  # Otherwise (missing RL3 file, and the corresponding <RUN_LEVEL> file):
  # Throw an error, letting users know how to create the file.
  stop(paste0("Missing file: `model3/", rl_dir, 'h3_MLE_eval.rds`;\nThe file can be obtained by running the script `H3_MLE_Eval.R`.'))
}
@

@@

@@

\section*{Materials and methods}
\subsection*{Mechanistic models for disease modeling}

@@

\new{In this article, we refer to} models that focus on learning relationships between variables in a dataset \new{as} \old{are called} {\it associative}, whereas models that incorporate a known scientific property of the system \old{are called} \new{we call} {\it causal} or {\it mechanistic}.
The danger in using forecasting techniques which rely on associative models to predict the consequence of interventions is called the Lucas critique in an econometric context.
Lucas et al.~\cite{lucas76} pointed out that it is naive to predict the effects of an intervention on a given system based entirely on historical associations.
To successfully predict the effect of an intervention, a model should therefore both provide a quantitative explanation of existing data and should have a causal interpretation: a manipulation of the system should correspond quantitatively with the corresponding change to the model.
This motivates the development of mechanistic models, which provides a statistical fit to the available data while also supporting a causal interpretation.
\new{Despite their limited ability to project the effect of interventions on a system, associative models can be effectively used to make inference on an certain features of a system.
In our literature review, there were 14 studies that used associative models to describe various aspects of the cholera epidemic \cite{hulland19,moise20,piarroux11,matias17,eisenberg13,rebaudet19CATI,ivers15,raila17,cuneo17,charles14,richterman19,richterman19inverse,bengtsson15,li16adjusting}.}

@@

We implemented Model~1 using the \code{pomp} package \cite{king16}, relying heavily on the \code{pomp} source code provided by Lee et al.~\cite{lee20}.
Both analyses used the \code{mif2} implementation of the IF2 algorithm to estimate $\paramVec$ by maximum likelihood.
One change we made in the statistical analysis that led to larger model likelihoods was increasing the computational effort in the numerical maximization.
While IF2 enables parameter estimation for a large class of models, the theoretic ability to maximize the likelihood depends on asymptotics in both the number of particles and the number of filtering iterations.
Many Monte Carlo replications are then required to quantify and further reduce the error.
The large increase in the log-likelihood for Model~1 (see Table~\ref{tab:likes}) can primarily be attributed to increasing the computational effort in fitting the model.
This result highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
\TODO{Possibly add more citations to the diagnostics part, maybe li23, laneri10 (Forcing Versus Feedback: Epidemic Malaria and Monsoon Rains in Northwest India)}
\new{In this case study, this was done by performing standard diagnostics for the IF2 and particle filter algorithms \cite{king16}. Given the considerable computational costs of simulation-based algorithms, we find it useful to perform an initial assessment using hyperparameter values that enable relatively quick calculations. The insights obtained from this preliminary analysis help in accurately determining the amount of computation that is required to achieve reliable outcomes.}
Simulations from the initial conditions of our fitted model are plotted against the observed incidence data in Fig.~\ref{fig:mod1fit}.

@@

Another approach is to compare a quantitative measure of the model fit (such as MSE, predictive accuracy, or model likelihood) among all proposed models.
These comparisons, which provide insight into how each model performs relative to the others, are quite common \cite{rinaldo12,sallah17}.
To calibrate relative measures of fit, it is useful to compare against a model that has well-understood statistical ability to fit data, and we call this model a {\it benchmark}.
Standard statistical models, interpreted as associative models without requiring any mechanistic interpretation of their parameters, provide suitable benchmarks.
Examples include linear regression, auto regressive moving average time series models, or even independent and identically distributed measurements.
Benchmarks enable us to evaluate the goodness of fit that can be expected of a suitable mechanistic model.

@@

\new{\editDiscussionSix}
